---
output:
  pdf_document: default
  html_document: default
---
# MSci Project

We start off by simulating trajectories using this model:

$Y_{ij} = a + b t_j + c t_j^2 + \epsilon_{ij}$

For this first task, we assume one cluster / class, here $i,j$ represent the individual and time point respectively. This is just for simplicity, and we can extend this later on. What I have done to begin with is:

Create the mean trajectory using Muntoni et al.

```{r}
# point 1 -  we start from age 1 and end at age 16
# from the graph we have that the average passes through points:
# (4, 22), (8,24) and (12,12)
# We get a = 6, b = 5.75, c = -0.4375

m = function(x){
  -0.4375*x^2 + 5.75*x + 6
}

# point 2, let's do 10 individual trajectories
generateValues <- function() {
  m_values <- m(1:16)
  rnorm_values <- rnorm(16, mean = 0, sd = 4) # Adjust the standard deviation (sd) here
  return(m_values + rnorm_values)
}

# create list of vectors
NSAA <- lapply(1:10, function(i) generateValues())

# create an AR(1) error correlation
# Set the parameters for the AR(1) error correlation
phi_error <- 0.9  # Autoregressive coefficient for the error
sigma_error <- 2.5 # Standard deviation of the error
errors_list <- vector("list", length = 10)

for (i in 1:10) {
  errors <- numeric(16)
  NSAA_i = vector(length = 16)
  error_t_minus_1 <- rnorm(1, mean = 0, sd = sigma_error)  # Initialize the error term
  for (j in 1:16) {
    val = min(max(m(j) + error_t_minus_1, 0), 34) 
    errors[j] = val - m(j)
    NSAA_i[j] = val
    error_t <- phi_error * error_t_minus_1 + rnorm(1, mean = 0, sd = sigma_error)  # Update the error term
    error_t_minus_1 <- error_t
  }
  NSAA[[i]] = NSAA_i
  errors_list[[i]] <- errors
}

all_errors <- unlist(errors_list)
pacf_errors <- pacf(all_errors, lag.max = 15, plot = FALSE)

# Plot the individual trajectories with AR(1) error correlation
png(filename = "plots/simulated_trajectories.png", width = 1000, height = 800, res = 150)
plot(1:16, NSAA[[1]], type = 'l', ylim = c(min(unlist(NSAA)), 34), xlab = 'Age', ylab = 'NSAA Total', main = 'Simulated Trajectories of 10 Patients')
for (i in 2:10) {
  lines(1:16, NSAA[[i]], type = 'l', col = i)
}
lines(1:16, m(1:16), col = "black", lwd = 2, lty = "dashed")
legend("topright", legend = c("Mean Trajectory"), col = "black", lwd = 2, lty = "dashed", title = NULL)
dev.off()

# Plot PACF for the AR(1) errors
png(filename = "plots/pacf_errors.png", width = 1000, height = 800, res = 150)
plot(pacf_errors, xlab = "Lag", ylab = "PACF", main = "Partial Autocorrelation of Errors")
abline(h = 0, col = "red")
dev.off()
# autocorrelation gives us some indication on how easily we can distinguish between whether the individual has been given treatment or not
```

I will now perform a hypothesis test on whether we can make the assumption that the final NSAA score from the "Natural History of Steroid-Treated Young Boys With Duchenne Muscular Dystrophy Using the NSAA" paper is equal to the sum of the baseline and the change over 12 months. All data used can be found in that paper.

```{r}
# age = 3 years
# test statistic = 19.0 - 17.3 = 2.7
# desired distribution is a normal with mean 3.8 and s.d. 5.6
# H0: mu = 3.8 H1: mu != 3.8 (two-tailed test)
# i.e. we are testing to see if the change between baseline 4 and 3 follows the same distribution as the change in NSAA over a 12 month period.
# find the S^2  = (n_3 * S_3^2 + n_4 * S_4^2 + ...) / sum of n_i's
# do hypothesis test for each age (comparing means) using S^2 divided by the sample sizes


# n_3 = 18, n_4 = 40, n_5 = 35, n_6 = 35, n_7 = 30 (from article)
# S_3.= 4.7^2, ... also from Natural History Article
S_squared = (18 * 4.7^2 + 40 * 4.4^2 + 35 * 3.7^2
             + 35 * 6.6^2 + 30 * 5.5^2) / (18 + 40
                                           + 35 + 35 + 30)

mean_change_3 = 3.8
var_mean_change_3 = 5.6
base_4 = 19.0
base_3 = 17.3
base_change_34 = base_4 - base_3
var_base_change_34 = S_squared / (40 + 18)
alpha = 0.05 # significance level
 
p = dnorm(mean_change_3 - base_change_34, mean=0, 
          sd=sqrt(var_mean_change_3 + var_base_change_34))
if(p<alpha / 2){
  print("observed value is significant - enough evidence to reject H0.")
} else{
print("Not enough evidence to reject H0 at this significance level.")
}
print(p)

# then do weighted regression between the mean change and the base change.
# create data frame with all the data points from the Natural History Article

df <- data.frame(age=c(3,4,5,6), base_change=c(1.7, 2.7, 1.4, -0.6),
                mean_change=c(3.8, 3.2, 0.5, -0.5))

# define weights
sizes = c(6, 20 ,22, 15)
ws = sizes

NSAA_lm <- lm(mean_change~base_change, data = df, weights = ws)
summary(NSAA_lm)
plot(NSAA_lm[2]) #QQ-plot of standardized residuals

# QQ-plot shows that standardized residuals roughly follow a N(0,1) distribution
# Indicates a good model, hence we can estimate mean_change by the base_change
```
