---
title: "extended_analysis"
author: "Vignesh Balaji"
date: "2024-10-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(tidyverse)
library(FactoMineR)
library(dplyr)
```

# Extended PCA

Here we attempt to illustrate a proof of concept for a new one-dimensional scoring system based off the PC1 of the chosen individual items. The idea is that we shall inspect the barycentric plots of each item and remove items which play very little impact in differentiating between patients, as exemplified through their temporal mean-variance relationship.

## Baseline Analysis

Read in baseline data
```{r}
baseline_scores <- read.csv("Data/baseline_scores.csv")
```

From visualising the plots, we can identify problematic variables as:

\item Get to Sitting
\item Rise from Floor
\item Stand
\item Stand on Heels
\item Walk

There are also some features that we should keep an eye on:

\item Hop Left and Right Legs
\item Stand up from Chair
\item Stand on one Leg - Left & Right

We should also consider whether we need assymmetric items - experiment with removing left leg activities and assume that right leg encodes enough information by itself.

```{r}

```
